---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

There are a few outliers that skew the data and would likely cause problems for analysis, such as Michelle Barrans's $251,258 salary or the 11 employees listed with a salary of $0. A detail that immediately jumped out to me was that most, or at least many, employees make six figures. I also noticed that many people have the same salary even at different positions.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
summarize(wh_salaries, mean_salary = mean(salary))
```

```{r}
summarize(wh_salaries, median_salary = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
summarize(wh_salaries, min_salary = min(salary))
```

```{r}
summarize(wh_salaries, max_salary = max(salary))
```


### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The median salary being slightly higher than the mean salary tells me the data is negatively skewed, likely because of the employees/detailees with salaries of $0. This is also evidenced in the minimum salary. The maximum salary in the dataset is also an outlier. The highest salary is almost $60,000 more than the second-highest salary.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

After analyzing the data, I believe the best story idea would be to investigate the statistical outliers, both at the bottom and top. I'd like to discover why 11 employees have a listed salary of $0 and explore what their roles entail, as well as why Michelle Barrans has such a high salary compared to the rest of the White House staff. My next steps toward developing this story would be to conduct online research into these employees and their positions. If I needed more information, I would contact or file a FOIA request for the Executive Office of the President/White House because they are responsible publishing this information.
