---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = (violent2019/pop2019 * 1000), property_rate_2019 = (property2019/pop2019 * 1000), violent_rate_2020 = (violent2020/pop2020 * 1000), property_rate_2020 = (property2020/pop2020 *1000))
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Baltimore and Ocean City remained atop the list of highest violent crime rates, though they both saw considerable changes between 2019 and 2020. Ocean City increased from about 13 to roughly 23 violent crimes per 1000 people, while Baltimore's violent crime rate decreased from about 18 to 15. Cambridge and Princess Anne also took sizable jumps into the top four in 2020. Ocean City climbed drastically in property crime rate as well. Ocean City already led in property crime rate, having 30 property crimes per 1000 people than the second-most, but in 2020, that disparity grew to roughly 65 more than the second-highest.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(violent_pct_change = ((violent_rate_2020 - violent_rate_2019)/violent_rate_2019 * 100), property_pct_change = ((property_rate_2020 - property_rate_2019)/property_rate_2019 * 100))
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

Some of the outliers may be noteworthy, such as a considerable increase from a crime rate of 0 or decrease to a rate of 0, but most of these take place in cities with quite low populations. The most newsworthy story in the data would be the large percentage changes in cities with substantial populations, such as Easton, Gaithersburg, Greenbelt and Bowie. All four of these cities' violent crime rates markedly increased. It's also worth noting that some of the larger cities in Prince George's County had increases in crime rate.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Lede:
Ocean City had the highest property crime rate in 2019 by far. In 2020 that gap widened even further.

Nutgraf:
Ocean City - one of Maryland's most popular vacation destinations - had 103 property crimes per 1000 people in 2019. Baltimore was a distant second with 73 property crimes per 1000 people in the same year. The following year Ocean City's leading rate jumped to roughly 121, while Baltimore remained in second with 56 property crimes per 1000 people. Of the ten cities with the highest property crime rates in 2019, eight of them saw that number dip in 2020. Ocean City experienced a 17% increase between those two years.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Percentiles

```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile = ntile(violent_rate_2020, 100))
```

```{r}
md_cities_percentiles |> summarize(IQR = IQR(violent_rate_2020))
```

This data may slightly change my story because Ocean City is in the 100th percentile of violent crime rate in 2020. My angle was about the property crime rate in Ocean City, but this data would likely make me cover both violent and property crime rate because the city being in the 100th percentile is certainly noteworthy.
