# JOUR405: Statistics for Journalists
## Midterm Exam - Spring 2025

Name: Trevor Gomes

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences. Start by loading the tidyverse and any other libraries you think you might need.

```{r}
library(tidyverse)
```


## Part 1: Restaurant Health Inspections (15 points)

You want to understand how restaurants in Montgomery County are performing on health inspections. The first dataset contains restaurant health inspection scores for restaurants in Montgomery County. The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv` and complete these tasks:

```{r}
moco_inspections <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```


### Tasks:
1. Calculate the mean and standard deviation of compliance scores (5 points)
2. Create a histogram of the compliance scores with a vertical line showing the mean (5 points)
3. Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? (5 points).

Task 1:
```{r}
moco_inspections %>% summarize(mean = mean(compliance_score), (sd = sd(compliance_score)))

```
Task 2:
```{r}
moco_inspections %>% 
  ggplot() +
  geom_histogram(aes(x = compliance_score), binwidth = 2) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1)

```

Task 3:
The mean is quite high at about 96.3, meaning there were few outliers with low scores. The data is skewed to the right, showing that most scores were 100 or near 100. Based on this data, the story would be how many inspections resulted in perfect or near perfect compliance scores.

## Part 2: High School Athletics (25 points)

You are reporting a story about high school sports participation in Maryland and want to see if there are differences between boys and girls. The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and sex. Load the data from: `https://raw.githubusercontent.com/example/md_hs_sports_2024.csv` and complete these tasks:

### Tasks:
1. Calculate the correlation between boys' and girls' participation (5 points)
2. Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. (5 points)
3. Create a scatterplot showing this relationship, adding a line of best fit (5 points)
4. In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

```{r}
hs_sports <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
```

Task 1:
```{r}
hs_sports %>% summarize(correlation = cor(boys, girls, method="pearson"))
```

Task 2:
```{r}
total <- hs_sports %>% mutate(total = (`girls` + `boys`), girls_pct = (`girls`/`total`)) 
```

Task 3:
```{r}
total %>% 
  ggplot() +
  geom_point(aes(x = girls, y = boys)) +
  geom_smooth(aes(x = girls, y = boys), method = "lm")
```

Task 4:
The correlation coefficient being 98, near a perfect positive correlation at 100, and the line of best fit sloping from left to right indicates that boys and girls participate in Motgomery County high school sports at similar rates. Schools above or below the line of best fit have disproportionate rates of boys or girls participating in sports compared to the opposite sex, and thus should be examined further because they oppose the trend of the data. A large majority of schools have roughly equal participation rates, but these outlier schools do not.

## Part 3: Public Transit Ridership (20 points)

You are investigating public transit ridership in the Washington, D.C. area and want to understand the patterns of daily bus and rail ridership. The third dataset contains daily bus and rail ridership totals from WMATA for the past year. Load the data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv and do the following:

### Tasks:
1. Calculate the average bus and rail ridership and standard deviation using summarize() (5 points)
2. Using the process you used in class, take a random sample daily ridership numbers and calculate the sample means and deviations for bus and rail. The number in the sample is up to you, but explain why you chose what you did. Compare this to the stats you generated in step 1. (5 points)
3. Using group_by() and summarize(), calculate the means for bus and rail ridership for each weekday. Describe the overall pattern of ridership for bus and rail - which days stand out and why? Are there differences between bus and rail in the standard deviation values? (10 points)


```{r}
dc_ridership <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```
Task 1:
```{r}
dc_ridership %>% summarize(mean = mean(bus), mean(rail), sd = sd(bus), sd(rail))
```

Task 2:
```{r}
sample_100 <- dc_ridership %>% 
  sample_n(100)
```

```{r}
sample_100 %>% summarize(mean = mean(bus), mean(rail), sd = sd(bus), sd(rail))
```
## I chose 100 as my sample size because the dataset is quite expansive, covering ridership data for every day in multiple years. 100 days is a good portion of a year, so I figured it would be representative of the population.

Task 3:
```{r}
dc_ridership %>% 
  group_by(weekday) %>% 
  summarize(mean = mean())
  
```


## Part 4: Maryland Car Theft Rates (20 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:

### Tasks:
1. Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate (5 points)
2. Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median, and what percentage of all car thefts occur in those counties? (5 points)
3. Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? (10 points)


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

Task 1:
```{r}
car_theft_rate <- car_thefts %>% mutate(car_theft_rate = ((`2023`/population)*250000))
```

Task 2:
```{r}
car_theft_rate %>% summarize(median = median(car_theft_rate))
```

```{r}
car_theft_rate %>% summarize(total = sum(`2023`))
```

```{r}
car_theft_rate %>% summarize(theft_percentage = (7138 + 8546 + 105 + 2427 + 92 + 2317 + 697 + 1080 + 262 + 276 + 150 + 106)/23871)
```


Task 3:

The counties with car theft rates above the median comprise 97 percent of car thefts statewide. The lede would be "12 Maryland counties make up 97 percent of the state's total car thefts."

## Part 5: Data Analysis Scenario (20 points)

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences (no code!) explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)

I would first obtain central tendencies - mean, median, and mode - of last year compared to this year. I would then calculate the percentage change from the mean (average) response time from last year to this year. This would capture how much the response time has worsened year over year in a single number.

2. What visualizations would help readers understand the trends? (5 points)

A line graph would most likely be the best option here. A line graph is best for displaying change over time, so the line could display how response times have changed in the past year. The x-axis could display months while the y-axis could display average response times.

3. What additional context or data would you need to make this a complete story? (5 points)

I would need to investigate potential changes in the police and fire departments to determine if there are any confounding variables. If a considerable number of employees were fired, this would obviously affect response times. I would also investiagte any potential changes in methods of response, in the way they handle phone calls or otherwise.

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Before submitting, clear your environment and run the entire notebook

Good luck!
