---
title: "HW9_TestScores"
name: Trevor Gomes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Did a New Reading Program Lead to Better Scores?

The superintendent recently claimed that a new reading program has improved third-grade reading scores across the school district.

Before the program, third-grade students in the district averaged 72.6 points on standardized reading tests with a standard deviation of 4.8 points.

After implementing the program for one semester, you collected scores from 12 randomly selected classrooms:
74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75

As a journalist, you need to determine: **Is there statistical evidence that reading scores have actually improved?**

## Task 1: Organize your data and initial assessment

Before you can run this codeblock, you will need to fill in a value where it says REPLACE_ME. That value can be found in the introduction.

```{r}
# Known information about reading scores before the new program
prior_mean <- 72.6  # average score
prior_sd <- 4.8     # standard deviation

# Reading scores after implementing the new program (12 classrooms)
new_scores <- c(74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75) # Replace with the actual scores

# Create a journalist-friendly dataset
score_data <- tibble(
  classroom = paste("Classroom", 1:12),
  reading_score = new_scores
)

# View the data
score_data
```

### Reflection Question 1:
Based on just looking at the score_data dataframe, have test scores improved? How can you tell?

The 12-score sample does suggest that test scores have improved after implementation of the new program because all 12 scores in the sample are higher than the previous average of 72.6.

## Task 2: Calculate key statistics

Like Task 1, you will need to replace values where it says REPLACE_ME before running any code.


```{r}
# Calculate statistics based on the new reading scores
new_stats <- score_data |> 
  summarise(
    mean = mean(reading_score),
    sd = sd(reading_score),
    n = n()
  )

new_stats
```

### Reflection Question 2:
Looking at the mean and standard deviation of the new scores compared to the previous statistics, what initial observations can you make? What questions might these statistics raise for your reporting?

The new mean derived from the 12-score sample, 75.75, is higher than the previous mean of 72.6. The standard deviation, though, is markedly lower likely because the sample is much smaller than the population. The low standard deviation raises the question of how large the population is, and thus, how accurately reflective the sample is. The higher mean also prompts me to question if there's a possible confounding variable causing the higher scores.

## Task 3: Create a column chart

As before, replace any values marked REPLACE_ME based the instructions.


```{r}
# STUDENT TASK: Choose an appropriate fill color for the bars
my_fill_color <- "lightblue" # Replace with a color name like "royalblue", "darkgreen", etc.

# Create a visualization comparing new scores to the previous average
score_data |> 
ggplot(aes(x = classroom, y = reading_score)) +
  geom_col(fill = my_fill_color, alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 2, y = prior_mean - 1, 
           label = "Previous Average (72.6)", hjust = 0, fontface = "bold", color = "darkred") +
  labs(
    title = "Reading Scores After New Program Implementation",
    subtitle = "Horizontal line shows previous district average of 72.6 points",
    x = NULL,
    y = "Reading Test Score",
    caption = "Source: District Assessment Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Reflection Question 3:
Examine the chart you created, and suggest a better title based on the results of the data, not a description.

"Data suggests reading scores have increased following new program implementation"

## Task 4: Perform a hypothesis test

This is where we formally test the superintendent's claim that reading scores have improved. Fill in the REPLACE_ME values as needed, beginning with your hypotheses.

**Hypotheses:**
Null: The new reading program has not led to improved reading scores
Alternative: The new reading program has led to improved reading scores

```{r}
# Set the significance level for your test
alpha_level <- 0.05 # Replace with the appropriate value

# Perform a one-sample t-test
# Since we want to know if scores improved (increased), we use a one-sided test (alternative = "greater")
t_test_result <- t.test(
  score_data$reading_score,
  mu = prior_mean,
  alternative = "greater"
)

# Display the results
t_test_result
```

### Reflection Question 4:
What does the p-value tell you, and what doesn't it tell you? How would you explain these results to a non-technical audience while maintaining accuracy?

A p-value as low as the one yielded by this t-test indicates "statistical significance" and a low likelihood that we could collect the 12 sample scores if the null hypothesis were true. This essentially means that there's a likely connection between the two variables - in our case, reading test scores and the implementation of the new reading program. A p-value does not tell you that a null hypothesis is certainly true or false, only how likely or unlikely obtaining a certain data set would be under the assumption that the null hypothesis is true. I would explain these results to a non-technical audience by saying that they indicate the reading program has likely improved reading test scores.

## Task 5: Interpreting the results for your news story

Let's gather all of the important stats we'll need in one place, so we can look at the prior average, the new scores and the results of the t.test, including the confidence interval. Replace any values where it says REPLACE_ME.


```{r}
# Get the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval
ci <- t.test(score_data$reading_score)$conf.int

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = prior_mean,
  `New average` = mean(new_scores),
  `Improvement` = mean(new_scores) - prior_mean,
  `Percent change` = round(((mean(new_scores) - prior_mean) / prior_mean) * 100, 1),
  `p-value` = p_value,
  `Lower bound` = ci[1],
  `Upper bound` = ci[2],
  `Confidence level` = "95%"
)

# Display the key statistics
story_stats
```

## Conclusion

### Reflection Question 5:
Based on these statistics, what would be your headline and lead paragraph for this story? Is there evidence to support the superintendent's claim?

HL: "Data suggests new reading program has improved third-grade reading scores across school district"

Third-grade reading scores increased by 4.16% district wide after the implementation of a new reading program. An analysis of 12 test scores after the reading program was instated supports the superintendent's claim that the new program has led to higher scores.

### Reflection Question 6:
What metrics or outcomes beyond test scores might be important to track for assessing reading performance?

Grades in English classes may be a considerable gauge of reading performance, as well as metrics tracking fluency or comprehension.
