
#Introduction/laying out what I'm investigating#  

In this project, I aim to determine if the duration of songs has significantly changed over time. First, I will analyze a dataset of roughly 160,000 songs retrieved from the Spotify API. The release date of these tracks ranges from 1921 to 2020. 

Then, I will narrow the dataset down to a specific decade, analyzing songs released from 2010 to 2020. Ultimately, my goal is to determine if the rise in prominence of social media, coinciding with a shortened attention span, has caused artists to decrease the duration of their songs. 

My null hypothesis is that there was no considerable change in song length during the 2010s. My alternate hypothesis is that there was considerable change in song length during the 2010s.
  
# Loading the tidyverse and my dataset

```{r}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(readr)
data <- read_csv("https://raw.githubusercontent.com/TrevorGomes/jour405_spring25/refs/heads/main/project/data.csv")
View(data)
```

# Here I convert milliseconds to seconds simply by dividing the duration_ms column by 1000. This makes the data more pallatable and easier to grasp.

```{r}
data_w_seconds <- data %>% mutate(
  duration_sec = duration_ms/1000)
```

# Evaluation: This created a new dataset called "data_w_seconds" featuring a new column (duration_sec)


# In this step, I'm summarizing the data to gather descriptive statistics and central tendencies. This data allows me to calculate z-scores in the next step.

```{r}
summary(data_w_seconds)
```
# Evaluation: The most significant statistic to my hypotheses is the duration of the songs. As seen above, the songs vary quite widely in duration, evidenced by the disparity between the median and mean, and minimum and maximum.


# In this block of code, I'm creating another dataset with four additional columns: the mean song duration, median song duration, the standard deviation of the song durations (all of which are the same for every entry in the entire dataset) and z-scores for each individual song. The z-score is calculated by subtracting the mean song duration of the whole dataset from an individual song's duration, then dividing the result by the standard deviation of song durations.

```{r}
data_w_z_scores <- data_w_seconds %>% mutate(
  mean_duration = mean(duration_sec),
  median_duration = median(duration_sec),
  sd_duration = sd(duration_sec),
  z_score = (duration_sec - mean_duration) / sd_duration)
```

# Evaluation: The z-scores for individual songs range widely, extending to over 40, meaning that song's length is over 40 standard deviations above the mean song duration.


# In this step, I'm creating a scatterplot that visualizes the relationship between song length and year. Year is assigned to the x-axis and song duration to the y-axis. The red line is the "line of best fit," that represents the general trend of the data. This graph gives a clearer visual representation of how song length has changed form 1921, the earliest date in the dataset, to 2020, the most recent data in the dataset. 

```{r}
data_w_z_scores |>
  ggplot(aes(x = year, y = duration_sec)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "year", y = "duration_sec", title = "Scatterplot of Song Length over the Years")
```

# Evaluation: Though the massive number of datapoints somewhat obscures the direction of the scatterplot, the line of best fit indicates a slight increase in song length from 1921 to 2020. Though this indicates a change, the increase is gradual, meaning, on the surface, this data would suggest my null hypothesis more so than my alternate hypothesis.


# In this step, I'm creating a linear model and conducting an analysis to determine how much effect the year has on the duration of a song. This analysis yields coefficients which indicate how much year (my independent variable) affects song duration (my dependent variable). The p-value and f-statistic indicate how statistically significant the model is, and the adjusted R-squared value represents the percentage of variance in the dependent variable that the independent variable is responsible for.

```{r}
model <- lm(`duration_sec` ~ `year`, data = data_w_z_scores)

summary(model)
```
# Evaluation: While the p-value and f-statistic show that the model is statistically significant, meaning the findings are unlikely to be caused by pure chance, the adjusted R-squared value and year coefficient are quite low. A ~ .006 adjusted R-squared value indicates that less than 1% of the variance in song duration can be ascribed to the year the song was released.


# In this step, I'm creating a line graph as another way to visualize the data and show the trend of song duration over the years.

```{r}
data_w_z_scores %>%
  ggplot(aes(x = year, y = duration_sec)) +
  geom_line(linetype = "dashed") +
  geom_smooth(method = "lm")
  labs(x = "year", y = "duration_sec", title = "Line Graph of Song Length over the Years")

```

# Evaluation: This line grpah reflects similiar findings to the scatterplot. As indicated by the blue line, song duration gradually increased from 1921 to 2020.


# Since there appears to be no significant change in song duration from 1921 to 2020, I will narrow the scope of my analysis to the 2010s. I will create a new dataset only including songs released from 2010 to 2020. This decade will be more telling for my hypotheses - if social media has considerably affected song length - and could exhibit more significant changes than the entire dataset. 

```{r}
data_2010s <- data_w_z_scores %>% filter(year == 2010 | year == 2011 | year == 2012 | year == 2013 | year == 2014 | year == 2015 | year == 2016 | year == 2017 | year == 2018 | year == 2019 | year == 2020)
```


# As I did for the dataset ranging from 1921 to 2020, I'm creating a scatterplot with a line of best fit to visualize any changes in average song length during the 2010s. The scatterplot is entirely the same, it now applies to a more specific range of dates.

```{r}
data_2010s |>
  ggplot(aes(x = year, y = duration_sec)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "year", y = "duration_sec", title = "Scatterplot of Song Length over the Years")
```

# Evaluation: This scatterplot shows a more pronounced negative relationship between song duration and year of release, which would lend itself more to my alternate hypothesis. Still, further analysis needs to be conducted to quantify the findings and determine if I can reject my null hypothesis.


# I'm also repeating the linear model analysis that I did for the entire dataset. This will allow me to see if my findings are statistically significant, and if the independent variable (year) has a more profound affect on the dependent variable (song duration) during the 2010s than from 1921 to 2020.

```{r}
model <- lm(duration_sec ~ year, data = data_2010s)

summary(model)
```

# Evaluation: The p-value and f-statistic in this rendition of the linear model analysis similarly indicate that the model is statistically significant. Though the adjusted R-squared value and year coefficients are markedly larger, they still don't reflect a considerable relationship between song duration and year. The adjusted R-squared value indicates that year of release is accountable for only roughly 3% of the variance of song length.


# As another way to visualize the data, I'm creating a line graph exclusively for the 2010s data, just as I did for the entire dataset.

```{r}
data_2010s %>%
  ggplot(aes(x = year, y = duration_sec)) +
  geom_line(linetype = "dashed") +
  geom_smooth(method = "lm")
  labs(x = "year", y = "duration_sec", title = "Line Graph of Song Length during the 2010s")
```

# Evaluation: The line graph showcases a similar slightly negative relationship between year and song duration.


# In this step, I'm determining the correlation coefficient between year and song duration, which quantifies the relationship between the two variables on a scale of 0 to 1 or 0 to -1.
```{r}
selected_vars <- data_2010s %>% select(`duration_sec`, `year`)
cor(selected_vars)
```

# Evaluation: Year and song duration have a correlation coefficient of -.1895, which is relatively weak, as expected by the gradually sloping line of best fit and low adjusted R-squared value. This coefficient suggests a slight negative relationship between song duration and year, but not strong enough of a relationship to reject my null hypothesis.


#This correlation matrix visualizes the correlation ceofficient (relationship) between song duration and year, by plotting the variables in a matrix and representing the strength of their relationship with the size of a circle.

```{r}
corr_matrix <- cor(selected_vars)
corrplot(corr_matrix, method = "circle")
```

# Evaluation: As seen above, the circles at the intersection of "year" and "duration_sec" are quite small and tan, representing a weak negative relationship. The other circles represent a perfectly positive relationship and dwarf the circle showing the correlation between year and song duration.



#Final story pitch

Through the analyses I've conducted, namely the linear model analyses that support the trends from the visualizations, I failed to find a significant relationship between song duration and year of release. From both the dataset as a whole, 1921 to 2020, and the 2010s decade specifically. The line of best fit appeared to have a low slope, only gradually increasing from 1921 to 2020, and slightly decreasing from 2010 to 2020. This was supported by the small adjusted R-squared values and coefficients for both ranges of data, indicating that year had a negligible effect on song duration.

These findings are significant, however. Much has been made of "TikTok music" and how social media has shortened users' attention spans, and with it, the duration of songs. Some claim that music as a whole is now tailored for social media platforms, and songs have been shortened. Though my analyses in this project suggest that, in a pool of over 160,000 songs over nearly a century, and nearly 22,000 songs released from 2010 to 2020, no such change has occurred.

My story would use these analyses to refute that claim and further examine why the shortening song narrative exists. I could conduct analyses on a newer set of songs, perhaps from 2020 to modern-day 2025 or specifically on songs that have trended on TikTok. An issue I'd likely face when pursuing this story would be gathering a dataset of songs that were popular on TikTok. 

For this story, I would consult someone who has researched recent music to interview them about their thoughts or findings on song duration. I would also reach out to a psychologist or other professional who has expertise on attention spans to ask them about social media's effect. I'd also inquire about how shortened attention span interacts with music, and what songs become popular.
